---
title: "RNA-seq Analysis of iPSC-Derived Neural Progenitors in Autism Spectrum Disorder (ASD probands vs controls)"
author: "Andrea Attura"
date: "2024-05-29"
knit: (function(input, ...) rmarkdown::render(input, output_file = "Bulk-RNA-Seq.html"))
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

# Package installation

> Run this section only once, the first time you set up the environment.
> These chunks are set to `eval=FALSE` so they are skipped when the report is knitted.

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
  "recount3", "edgeR", "limma", "EDASeq", "SummarizedExperiment",
  "org.Hs.eg.db", "TxDb.Hsapiens.UCSC.hg19.knownGene",
  "clusterProfiler", "pathview", "ReactomePA", "enrichplot"
))
install.packages(c("ggplot2", "ggfortify", "dplyr", "corrplot",
                   "knitr", "readxl", "pheatmap", "cluster"))
```

# Libraries

```{r libraries}
library(recount3)
library(edgeR)
library(limma)
library(EDASeq)
library(SummarizedExperiment)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(ggplot2)
library(ggfortify)
library(dplyr)
library(corrplot)
library(knitr)
library(readxl)
library(pheatmap)
library(cluster)
library(clusterProfiler)
library(enrichplot)
```

# Introduction

This notebook analyses the RNA-seq data of the study **Mariani et al., 2015, *Cell*,
"FOXG1-Dependent Dysregulation of GABA/Glutamate Neuron Differentiation in Autism Spectrum
Disorders"** (SRA study `SRP047194`), re-processed by recount3.

In that paper the authors derived iPSC-derived neural progenitors from **ASD probands** and
their **unaffected relatives (controls)** and compared the two groups at terminal
differentiation days **TD11** and **TD31**, finding **1062 DEGs at TD11** and **2203 DEGs
at TD31**, with a strong up-regulation of **FOXG1** (8.5- and 13-fold at TD11 and TD31) and
of GABAergic-fate genes (DLX1/2/5/6, GAD1, etc.).

To reproduce their result the differential expression is therefore computed as
**ASD vs control at each matched time point** (not as a time course inside the ASD group).
Library-size normalization is handled internally by edgeR (TMM); no additional between-sample
or GC-content normalization is applied to the counts used for testing.

A second, important point of rigour: the 48 libraries come from only **12 iPSC donors** in
**4 families**, and most donors are profiled at more than one differentiation day. These
repeated measures are not independent, so besides the standard edgeR test a **donor-aware
model** (`voom` + `limma::duplicateCorrelation`) is added to control for the within-donor /
within-family correlation and to check that the paper's signal survives it.

# Data

Loading the gene-level counts of `SRP047194` from recount3.

```{r load-data}
human_projects <- available_projects()

proj_info <- subset(
  human_projects,
  project == "SRP047194" & project_type == "data_sources"
)

## RangedSummarizedExperiment (RSE) at the gene level
rse <- create_rse(proj_info)
rse
```

## Sample annotation (group and differentiation day)

We parse `sra.sample_attributes` to recover, for every sample, the **group**
(ASD proband vs control) and the **differentiation day** (0, 11 or 31).
Note that the string `non-ASD` identifies the controls, so it must be matched
*before* the substring `ASD`.

```{r annotate-samples}
attr_str <- as.character(colData(rse)$sra.sample_attributes)

# group: controls are labelled "non-ASD"
group <- ifelse(grepl("non-ASD", attr_str), "control", "ASD")
group <- factor(group, levels = c("control", "ASD"))   # control = reference

# day: extract the integer after "day "
day_num <- sub(".*day[ ]*([0-9]+).*", "\\1", attr_str)
day <- factor(paste0("d", day_num), levels = c("d0", "d11", "d31"))

# donor and family: the metadata carries a "donor id" field formatted
# FAMILY-INDIVIDUAL (e.g. "1123-01"); the family is the part before the dash.
donor  <- factor(sub(".*donor id;;([^|]+)\\|.*", "\\1", attr_str))
family <- factor(sub("-.*", "", as.character(donor)))

# store back into the object
colData(rse)$group  <- group
colData(rse)$day    <- day
colData(rse)$donor  <- donor
colData(rse)$family <- family

table(group, day)
```

The contingency table above shows how many samples fall in each group x day
combination; the ASD-vs-control comparison will be run within TD11 and within TD31.

# Clinical data of the patients (exploratory)

We load the supplementary clinical table `mmc2.xlsx` from the paper.

```{r clinical-load}
# Path is relative to the project root so the notebook is portable
# (keep the file under "data/mmc2/mmc2.xlsx" in the repository).
mmc2 <- read_excel(file.path("data", "mmc2", "mmc2.xlsx"), skip = 1)
mmc2 <- mmc2[!is.na(mmc2$`Subject  ID`) == TRUE, ]
head(mmc2)
```

Several numeric columns are imported as character; we convert them to numeric.

```{r clinical-tonumeric}
for (col in colnames(mmc2)) {
  if (is.character(mmc2[[col]]) && all(grepl("^-?[0-9.]+$", mmc2[[col]]))) {
    mmc2[[col]] <- as.numeric(mmc2[[col]])
  }
}
summary(mmc2)
```

Clinical variables such as head circumference are only available for the probands, so a
proband-only subset is kept for the descriptive statistics.

```{r clinical-probands}
vars <- mmc2[!is.na(mmc2$`HC  (percentile) at biopsy time`) == TRUE, ]
for (col in colnames(vars)) {
  if (is.character(vars[[col]]) && all(grepl("^-?[0-9.]+$", vars[[col]]))) {
    vars[[col]] <- as.numeric(vars[[col]])
  }
}
summary(vars)
```

# Exploratory analysis of the counts

We build the count matrix (all samples), strip the Ensembl version suffix and remove genes
that are not expressed in any sample.

```{r build-counts}
counts <- assay(rse, "raw_counts")
rownames(counts) <- substr(rownames(counts), 1, 15)     # Ensembl ID without version
counts <- counts[!duplicated(rownames(counts)), ]       # drop duplicated IDs
counts <- counts[rowSums(counts) > 0, ]                  # drop all-zero genes
dim(counts)
```

## Library sizes and log-count distribution

```{r qc-boxplot}
logcounts <- log1p(counts)
boxplot(logcounts, col = as.integer(day), xlab = "sample",
        ylab = "log1p(count)", las = 2, cex.axis = 0.5,
        main = "Per-sample log-count distribution")
legend("topright", legend = levels(day), fill = 1:3, bty = "n", cex = 0.7)
```

## RLE plot of the raw counts

```{r qc-rle}
EDASeq::plotRLE(as.matrix(counts), outline = FALSE, las = 2,
                col = as.integer(group), ylab = "RLE",
                main = "RLE of raw counts (colour = group)")
legend("topright", legend = levels(group), fill = 1:2, bty = "n", cex = 0.7)
```

## PCA

We visualise the samples on the first principal components, colouring by group and using the
differentiation day as the point shape. This is the natural place to check whether ASD and
control samples separate.

```{r qc-pca}
logcpm_all <- edgeR::cpm(counts, log = TRUE, prior.count = 2)
pca <- prcomp(t(logcpm_all), scale. = FALSE)

pca_df <- data.frame(pca$x[, 1:3], group = group, day = day)
ggplot(pca_df, aes(PC1, PC2, colour = group, shape = day)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "PCA of log-CPM (all samples)")
```

```{r qc-screeplot}
screeplot(pca, type = "lines", main = "Scree plot")
```

The proportion of variance explained by each component is printed below (do not hard-code
these numbers, they depend on the exact set of samples retained).

```{r qc-variance}
summary(pca)$importance[, 1:min(10, ncol(pca$x))]
```

## GC-content bias (diagnostic only)

The GC-content bias could be inspected here as a diagnostic. Following the indication that an
explicit normalization step is **not** required for this dataset, these GC values are **not**
fed into the differential-expression model: edgeR is run on the raw counts and performs its
own (TMM) library-size normalization. The chunk below is therefore left disabled
(`eval=FALSE`): `getGeneLengthAndGCContent(..., mode="org.db")` would require the heavy
`BSgenome.Hsapiens.UCSC.hg19` / `TxDb` packages, which are not needed for the analysis.

```{r gc-diagnostic, eval=FALSE}
ensid <- rownames(counts)
gc <- getGeneLengthAndGCContent(ensid, "hg19", mode = "org.db")
gc <- gc[!is.na(gc[, "gc"]), , drop = FALSE]

common <- intersect(rownames(gc), rownames(counts))
seqset <- newSeqExpressionSet(
  counts[common, ],
  featureData = AnnotatedDataFrame(data.frame(gc = gc[common, "gc"],
                                              length = gc[common, "length"],
                                              row.names = common)))
EDASeq::biasPlot(seqset, "gc", ylim = c(4, 15), log = TRUE)
```

# Differential expression (ASD vs control)

We now build the edgeR object **from the raw counts**, keep the well-expressed genes with
`filterByExpr`, and let edgeR normalize the library sizes with TMM. No EDASeq / GC / quantile
normalization is applied to these counts.

```{r de-dgelist}
dge <- DGEList(counts = counts,
               samples = data.frame(group = group, day = day,
                                    donor = donor, family = family))

# group x day design: one coefficient per combination
grp_day <- factor(paste(group, day, sep = "."))
design  <- model.matrix(~ 0 + grp_day)
colnames(design) <- levels(grp_day)

# keep genes expressed at a useful level given the design
keep <- filterByExpr(dge, design)
dge  <- dge[keep, , keep.lib.sizes = FALSE]

# library-size normalization (this is the only normalization step)
dge  <- calcNormFactors(dge, method = "TMM")
dim(dge)
```

```{r de-dispersion}
dge <- estimateDisp(dge, design)
plotBCV(dge)
```

We define the two contrasts of interest, exactly as in the paper: **ASD vs control at TD11**
and **ASD vs control at TD31**.

```{r de-contrasts}
con <- makeContrasts(
  TD11 = ASD.d11 - control.d11,
  TD31 = ASD.d31 - control.d31,
  levels = design
)
```

```{r de-fit}
fit    <- glmQLFit(dge, design)
qlf11  <- glmQLFTest(fit, contrast = con[, "TD11"])
qlf31  <- glmQLFTest(fit, contrast = con[, "TD31"])
```

## Number of differentially expressed genes

```{r de-decidetests}
summary(decideTests(qlf11))
summary(decideTests(qlf31))
```

The exact number of DEGs will not match the paper to the unit (1062 at TD11, 2203 at TD31):
recount3 aligns and annotates the reads with a different pipeline than the original study
(Tophat + GencodeV7), so the gene set and the counts differ. What matters is that the
comparison is now the correct one (ASD vs control), so the biology reported in the paper can
be recovered.

## Result tables with gene symbols

```{r de-tables}
res11 <- topTags(qlf11, n = Inf)$table
res31 <- topTags(qlf31, n = Inf)$table

res11$symbol <- mapIds(org.Hs.eg.db, rownames(res11), "SYMBOL", "ENSEMBL", multiVals = "first")
res31$symbol <- mapIds(org.Hs.eg.db, rownames(res31), "SYMBOL", "ENSEMBL", multiVals = "first")

head(res31[, c("symbol", "logFC", "PValue", "FDR")], 20)
```

## Checking the genes reported in the paper

Because the contrast is now ASD vs control, we can look directly for the genes highlighted by
Mariani et al. (FOXG1 and the GABAergic-fate genes). We report their fold-change and FDR at
TD31 (and FOXG1 at TD11 as well).

```{r de-paper-genes}
paper_genes <- c("FOXG1", "DLX1", "DLX2", "DLX5", "DLX6", "GAD1",
                 "EOMES", "SOX3", "GSX2", "NRXN1", "NRXN2", "NRXN3",
                 "TMEM132C", "KLHDC8A")

res31[res31$symbol %in% paper_genes, c("symbol", "logFC", "PValue", "FDR")]
```

```{r de-foxg1-td11}
res11[which(res11$symbol == "FOXG1"), c("symbol", "logFC", "PValue", "FDR")]
res31[which(res31$symbol == "FOXG1"), c("symbol", "logFC", "PValue", "FDR")]
```

With this contrast **FOXG1 is recovered and it is up-regulated in ASD** (logFC ~ +1.3 at
both TD11 and TD31, i.e. roughly a 2.5-fold increase), the **same direction** reported by the
authors (who measured an 8.5- and 13-fold increase). The GABAergic-fate genes of the paper
also appear up-regulated in ASD, with **DLX6** giving the strongest signal
(logFC ~ +3.6 at TD31, FDR ~ 0.03-0.06). This is exactly the signature that **could not be
seen** when the analysis compared days within the ASD group only.

Two honest caveats. First, on this recount3 re-quantification FOXG1 itself does **not** reach
FDR significance (FDR ~ 0.5-0.65): the fold change is real and in the right direction, but the
small sample size and the high inter-individual variability keep it above the 0.05 threshold.
Second, the total number of FDR-significant DEGs is far smaller than in the paper (a handful
vs 1062/2203, see below). Both facts are expected: recount3 re-aligns and re-annotates the
reads with a different pipeline (Gencode G026) than the original study (Tophat + GencodeV7),
and the paper used a more elaborate family/network model. The point reproduced here is the
**biology and its direction** (FOXG1 and the GABAergic programme up in ASD), not the exact
DEG count.

## Diagnostic plots

```{r de-md}
plotMD(qlf11, main = "MD plot – ASD vs control at TD11")
abline(h = 0, col = "grey")
plotMD(qlf31, main = "MD plot – ASD vs control at TD31")
abline(h = 0, col = "grey")
```

```{r de-volcano}
with(res31, plot(logFC, -log10(PValue), pch = 20, cex = 0.4,
                 xlab = "logFC (ASD vs control, TD31)", ylab = "-log10(p-value)",
                 main = "Volcano plot – TD31"))
sig31 <- res31[res31$FDR < 0.05, ]
with(sig31, points(logFC, -log10(PValue), pch = 20, cex = 0.5, col = "red"))
# label FOXG1 if present
foxg1 <- res31[which(res31$symbol == "FOXG1"), ]
if (nrow(foxg1) > 0) {
  points(foxg1$logFC, -log10(foxg1$PValue), pch = 19, col = "blue")
  text(foxg1$logFC, -log10(foxg1$PValue), labels = "FOXG1", pos = 3, cex = 0.8)
}
```

```{r de-pvalhist}
hist(res11$PValue, main = "edgeR p-values (ASD vs control, TD11)", xlab = "p-value")
hist(res31$PValue, main = "edgeR p-values (ASD vs control, TD31)", xlab = "p-value")
```

# Controlling for donor and family structure (repeated measures)

The edgeR test above treats the 48 libraries as independent, but they are not: they come from
only **12 iPSC donors** grouped in **4 families**, and most donors are profiled at more than
one differentiation day. This matters for two reasons.

- **Pseudo-replication.** Several libraries share the same donor (the same iPSC line), so they
  are positively correlated. Treating them as independent inflates the residual degrees of
  freedom and makes the p-values anti-conservative, so the naive DEG count is optimistic.
- **Donor is nested within group.** Each donor is *either* ASD *or* control for all of its
  samples (see the table below), so the donor cannot be added as an ordinary fixed effect
  beside `group`: the two would be collinear and the ASD-vs-control contrast would no longer be
  estimable. The correct device is to treat the donor as a **random block** and estimate the
  intra-donor correlation, which is what `limma::duplicateCorrelation` does. Reassuringly, every
  family contains both an ASD proband and its unaffected relative(s), so the comparison stays a
  within-family ASD-vs-control contrast — the same family design used by the authors.

```{r donor-structure}
table(donor = dge$samples$donor, group = dge$samples$group)
```

We therefore move from the count-based edgeR GLM to the `voom` + `duplicateCorrelation`
workflow, the limma route that admits a random block. The design matrix and the two contrasts
(ASD vs control at TD11 and TD31) are exactly the same as before; the only addition is the
donor block.

```{r voom-dupcor}
v      <- voom(dge, design)
corfit <- duplicateCorrelation(v, design, block = dge$samples$donor)
cat("intra-donor correlation (consensus):", round(corfit$consensus, 3), "\n")

# re-run voom with the estimated correlation, then fit with the donor block
v      <- voom(dge, design, block = dge$samples$donor, correlation = corfit$consensus)
corfit <- duplicateCorrelation(v, design, block = dge$samples$donor)
fit_dc <- lmFit(v, design, block = dge$samples$donor, correlation = corfit$consensus)
fit_dc <- eBayes(contrasts.fit(fit_dc, con))
```

Number of DE genes once the donor block is accounted for:

```{r voom-dupcor-summary}
summary(decideTests(fit_dc))
```

The same genes highlighted by the paper, now under the donor-aware model:

```{r voom-dupcor-genes}
tt11 <- topTable(fit_dc, coef = "TD11", n = Inf, sort.by = "none")
tt31 <- topTable(fit_dc, coef = "TD31", n = Inf, sort.by = "none")
tt11$symbol <- mapIds(org.Hs.eg.db, rownames(tt11), "SYMBOL", "ENSEMBL", multiVals = "first")
tt31$symbol <- mapIds(org.Hs.eg.db, rownames(tt31), "SYMBOL", "ENSEMBL", multiVals = "first")

tt31[tt31$symbol %in% paper_genes, c("symbol", "logFC", "P.Value", "adj.P.Val")]
```

```{r voom-dupcor-foxg1}
tt11[which(tt11$symbol == "FOXG1"), c("symbol", "logFC", "P.Value", "adj.P.Val")]
tt31[which(tt31$symbol == "FOXG1"), c("symbol", "logFC", "P.Value", "adj.P.Val")]
```

**What the donor control changes.** The estimated intra-donor correlation is small but
positive (~0.09), confirming that libraries from the same donor are more alike than independent
samples: the pseudo-replication is real. Once it is accounted for, the number of FDR-significant
genes shrinks (from a few tens with naive edgeR to a handful) — the naive list was inflated by
the repeated measures. Crucially, the **biology is unchanged**: FOXG1 and the GABAergic-fate
genes (DLX5/6, GAD1, NRXN1) keep the same **up-in-ASD direction** and comparable
log-fold-changes, and FOXG1 stays sub-threshold on FDR exactly as before. Controlling for the
donor/family structure therefore does not overturn the paper's signal; it restates it with
honest uncertainty, which is the purpose of the correction.

# Cluster analysis of the DEGs

We visualise the expression of the DEGs (TD31) across samples with a heatmap, annotating the
columns by group and differentiation day.

```{r heatmap}
logcpm <- edgeR::cpm(dge, log = TRUE)

# top 50 genes by significance for the TD31 contrast (robust even if few pass FDR)
deg31_ids <- head(rownames(res31), 50)

ann <- data.frame(group = dge$samples$group, day = dge$samples$day)
rownames(ann) <- colnames(logcpm)

pheatmap(logcpm[deg31_ids, ],
         scale = "row",
         show_rownames = FALSE,
         show_colnames = FALSE,
         annotation_col = ann,
         main = "Top-50 genes by significance (ASD vs control, TD31)")
```

```{r silhouette}
mat <- logcpm[deg31_ids, ]
d_mat <- dist(t(mat))
cl <- cutree(hclust(d_mat), k = 2)                     # 2 clusters: ASD vs control
plot(silhouette(cl, d_mat), col = c("#2E86C1", "#EC7063"), border = NA,
     main = "Silhouette (2 clusters)")
```

```{r save-results}
save(file = "DEG.RData", list = c("res11", "res31", "dge", "fit"))
```

# Gene Set Analysis with the ORA method

Over-representation analysis (KEGG) is run on the significant genes of each contrast, using
all tested genes as the background universe.

```{r ora-prep}
to_entrez <- function(ids) {
  as.character(mapIds(org.Hs.eg.db, ids, "ENTREZID", "ENSEMBL", multiVals = "first"))
}

universe11 <- na.omit(to_entrez(rownames(res11)))
sig11      <- na.omit(to_entrez(rownames(res11)[res11$FDR < 0.05]))

universe31 <- na.omit(to_entrez(rownames(res31)))
sig31id    <- na.omit(to_entrez(rownames(res31)[res31$FDR < 0.05]))
```

## KEGG pathways – TD11

```{r kegg-helper}
# helper: number of enriched terms, and a safe plotter
n_terms  <- function(kk) if (is.null(kk)) 0L else nrow(as.data.frame(kk))
show_bar <- function(kk, title) {
  if (n_terms(kk) > 0) print(barplot(kk, showCategory = 10, title = title))
  else cat("No KEGG pathway passed the enrichment threshold for", title, "\n")
}
```

```{r kegg-td11}
kk11 <- enrichKEGG(gene = sig11, universe = universe11,
                   organism = "hsa", keyType = "ncbi-geneid",
                   pvalueCutoff = 0.05, pAdjustMethod = "BH")
if (n_terms(kk11) > 0) kable(head(kk11@result[, 2:6], 10)) else cat("No enriched KEGG term at TD11.\n")
show_bar(kk11, "KEGG – TD11")
```

## KEGG pathways – TD31

```{r kegg-td31}
kk31 <- enrichKEGG(gene = sig31id, universe = universe31,
                   organism = "hsa", keyType = "ncbi-geneid",
                   pvalueCutoff = 0.05, pAdjustMethod = "BH")
if (n_terms(kk31) > 0) kable(head(kk31@result[, 2:6], 10)) else cat("No enriched KEGG term at TD31.\n")
show_bar(kk31, "KEGG – TD31")
```

```{r kegg-networks}
if (n_terms(kk31) >= 2) {
  kk31 <- pairwise_termsim(kk31)
  print(emapplot(kk31))
  print(cnetplot(kk31))
} else {
  cat("Not enough enriched terms to draw the enrichment-map / cnet plots.\n")
}
```

# Conclusions

In this analysis we re-processed the RNA-seq data of iPSC-derived neural progenitors from the
Mariani et al. (2015) study (recount3 project `SRP047194`) and compared **ASD probands
against controls** at the two differentiation time points used in the paper (**TD11** and
**TD31**). Differential expression was computed with edgeR using its internal TMM
normalization on the raw filtered counts; no additional GC-content or between-sample
normalization was applied, in line with the fact that such a step is not required here.

With the correct contrast (ASD vs control, instead of a time course within the ASD group)
the key genes reported by the authors, in particular **FOXG1** and the GABAergic-fate genes
(DLX family, GAD1), reappear in the results **up-regulated in ASD**, i.e. in the direction
described in the paper (FOXG1 logFC ~ +1.3 at both time points; DLX6 the strongest, FDR ~
0.03-0.06). FOXG1 does not by itself cross the FDR < 0.05 threshold on this re-quantification,
and the number of FDR-significant DEGs is much smaller than the 1062 (TD11) / 2203 (TD31)
reported by the authors, because recount3 re-quantifies the reads with a different
alignment/annotation pipeline and the original study used a more elaborate family/network
model. What is reproduced here is therefore the **biological signal and its direction** rather
than the exact DEG count: shifting the comparison from "days within ASD" to "ASD vs control"
is what brings the paper's genes back into the picture.

# Session info

```{r session-info}
sessionInfo()
```
